# "Best Subsets" Variable Selection in our Prostate Cancer Study

A second approach to model selection involved fitting all possible subset models and identifying the ones that look best according to some meaningful criterion and ideally one that includes enough variables to model the response appropriately without including lots of redundant or unnecessary terms. 

Here's the set of predictors we're considering for our modeling of `lpsa`, our outcome. Note that five of the eight predictors are quantitative, and the remaining three (`bph`, `svi` and `gleason`) are categorical. A little cleaning gives us the following results.

```{r}
prost_c12 <- prost %>%
    mutate(bph = fct_relevel(factor(bph), "High", "Medium"),
           gleason = fct_relevel(factor(gleason), "6", "7"),
           svi = factor(svi))

prost_c12 %>% select(-subject, -lpsa) %>% summary()
```

## Key Summaries We'll Use to Evaluate Potential Models (in-sample)

1. Adjusted $R^2$, which we try to maximize.
2. Bayesian Information Criterion (BIC), which we also try to minimize.
3. Mallows' $C_p$ statistic, which we will try to minimize, and which is closely related to AIC in this setting.

Choosing between AIC and BIC can be challenging.

> For model selection purposes, there is no clear choice between AIC and BIC. Given a family of models, including the true model, the probability that BIC will select the correct model approaches one as the sample size n approaches infinity - thus BIC is asymptotically consistent, which AIC is not. [But, for practical purposes,] BIC often chooses models that are too simple [relative to AIC] because of its heavy penalty on complexity. 

- Source: @Hastie2001, page 208.

Several useful tools for running "all subsets" or "best subsets" regression comparisons are developed in R's `leaps` package.

## Using `regsubsets` in the `leaps` package

We can use the `leaps` package to obtain results in the `prost` study from looking at all possible subsets of the candidate predictors. The `leaps` package isn't particularly friendly to the tidyverse. In particular, we **cannot have any character variables** in our predictor set. We specify our "kitchen sink" model, and apply the `regsubsets` function from `leaps`, which identifies the set of models.

To start, we'll ask R to find the one best subset (with 1 predictor variable [in addition to the intercept], then with 2 predictors, and then with each of 3, 4, ... 8 predictor variables) according to an exhaustive search without forcing any of the variables to be in or out. 

- Use the `nvmax` command within the `regsubsets` function to limit the number of regression inputs to a maximum.
- Use the `nbest` command to identify how many subsets you want to identify for each predictor count.

```{r}
rs_mods <- regsubsets(
    lpsa ~ lcavol + lweight + age + bph + 
        svi + lcp + gleason + pgg45,
    data = prost_c12, nvmax = 8, nbest = 1)

rs_summ <- summary(rs_mods)
```

### Identifying the models with `which` and `outmat`

The summary of the regsubsets output provides lots of information we'll need. To see the models selected by the system, we use:


```{r}
rs_summ$which
```

Another version of this formatted for printing is:

```{r}
rs_summ$outmat
```

So...

- the best one-predictor model used `lcavol`
- the best two-predictor model used `lcavol` and `lweight`
- the best three-predictor model used `lcavol`, `lweight` and `svi`
- the best four-predictor model added an indicator variable for `gleason7`
- the best five-predictor model added an indicator for `bphLow` and `age` to model 3
- the best six-predictor model added a `gleason6` indicator to model 5
- the best seven-predictor model added `lcp` to model 6,
- and the eight-input model replaced `gleason6` with `gleason7` and also added `pgg45`.

These "best subsets" models are not hierarchical. If they were, it would mean that each model was a subset of the one below it. 

- To identify potentially attractive candidate models, we can either tabulate or plot key summaries of model fit (adjusted $R^2$, Mallows' $C_p$ and BIC) using `ggplot2`. I'll show both approaches.

### Obtaining Fit Quality Statistics

We'll first store the summary statistics and winning models at each input count into a tibble called `rs_winners`. We're doing this to facilitate plotting with `ggplot2`.

```{r}
rs_winners <- tbl_df(rs_summ$which) %>%
    mutate(inputs = 1:(rs_mods$nvmax - 1),
           r2 = rs_summ$rsq,
           adjr2 = rs_summ$adjr2,
           cp = rs_summ$cp,
           bic = rs_summ$bic,
           rss = rs_summ$rss) %>%
    select(inputs, adjr2, cp, bic, everything())

rs_winners
```

## Plotting the Fit Statistics

### Plots of $R^2$ and RMSE for each model

```{r}
ggplot(rs_winners, aes(x = inputs, y = r2,
                       label = round(r2,3))) +
    geom_line() +
    geom_label() +
    geom_label(data = subset(rs_winners,
                             r2 == max(r2)),
               aes(x = inputs, y = r2,
                   label = round(r2,3)),
               fill = "black", col = "white") +
    labs(x = "# of regression inputs",
         y = "R-squared")
```

Remember that raw $R^2$ is greedy, and will always look to select as large a set of predictors as possible. The residual sum of squares, if plotted, will have a similar problem, where the minimum RSS will always be associated with the largest model (if the models are subsets of one another.)

```{r}
ggplot(rs_winners, aes(x = inputs, y = rss,
                       label = round(rss,1))) +
    geom_line() +
    geom_label() +
    geom_label(data = subset(rs_winners,
                             rss == min(rss)),
               aes(x = inputs, y = rss,
                   label = round(rss,1)),
               fill = "black", col = "white") +
    labs(x = "# of regression inputs",
         y = "Residual Sum of Squares")
```

So the $R^2$ and residual sums of squares won't be of much help to us. Instead, we'll focus on the three other measures we mentioned earlier.

1. Adjusted $R^2$ which we try to maximize,
2. Mallows' $C_p$ which we try to minimize, and
3. The Bayes Information Criterion (BIC) which we also try to minimize.

### Adjusted $R^2$ values for our subsets

```{r}
p1 <- ggplot(rs_winners, aes(x = inputs, y = adjr2, 
                       label = round(adjr2,3))) +
    geom_line() +
    geom_label() +
    geom_label(data = subset(rs_winners, 
                             adjr2 == max(adjr2)),
               aes(x = inputs, y = adjr2, 
                   label = round(adjr2,3)), 
               fill = "yellow", col = "blue", size = 6) +
    scale_y_continuous(expand = expand_scale(mult = .1)) +
    labs(x = "# of regression inputs",
         y = "Adjusted R-squared")

p1
```

### Mallows' $C_p$ for our subsets

The $C_p$ statistic focuses directly on the tradeoff between **bias** (due to excluding important predictors from the model) and extra **variance** (due to including too many unimportant predictors in the model.) 

If N is the sample size, and we select $p$ regression predictors from a set of $K$ (where $p < K$), then the $C_p$ statistic is

$C_p = \frac{SSE_p}{MSE_K} - N + 2p$

where:

- $SSE_p$ is the sum of squares for error (residual) in the model with $p$ predictors
- $MSE_K$ is the residual mean square after regression in the model with all $K$ predictors

As it turns out, this is just measuring the particular model's lack of fit, and then adding a penalty for the number of terms in the model (specifically $2p - N$ is the penalty since the lack of fit is measured as $(N-p) \frac{SSE_p}{MSE_K}$.

- If a model has no meaningful lack of fit (i.e. no substantial bias) then the expected value of $C_p$ is roughly $p$. 
- Otherwise, the expectation is $p$ plus a positive bias term. 
- In general, we want to see *smaller* values of $C_p$. 

```{r}
p2 <- ggplot(rs_winners, aes(x = inputs, y = cp, 
                       label = round(cp,1))) +
    geom_line() +
    geom_label() +
    geom_label(data = subset(rs_winners, 
                             cp == min(cp)),
               aes(x = inputs, y = cp, 
                   label = round(cp,1)), 
               fill = "navy", col = "white", size = 6) +
    scale_y_continuous(expand = expand_scale(mult = .1)) +
    labs(x = "# of regression inputs",
         y = "Mallows' Cp")

p2
```

### BIC for our subsets

We might consider several information criteria:

- the Bayesian Information Criterion, called BIC
- the Akaike Information Criterion (used by R's default stepwise approaches,) called AIC
- a corrected version of AIC due to @HurvichTsai1989, called $AIC_c$ or `aic.c`

Each of these indicates better models by getting smaller. Since the $C_p$ and AIC results will lead to the same model, I'll focus on plotting the BIC.

```{r}
p3 <- ggplot(rs_winners, aes(x = inputs, y = bic, 
                       label = round(bic, 1))) +
    geom_line() +
    geom_label() +
    geom_label(data = subset(rs_winners, 
                             bic == min(bic)),
               aes(x = inputs, y = bic, label = round(bic,1)), 
               fill = "red", col = "white", size = 6) +
    scale_y_continuous(expand = expand_scale(mult = .1)) +
    labs(x = "# of regression inputs",
         y = "Bayes Information Criterion")

p3
```

We could, if necessary, also calculate the *uncorrected* AIC value for each model, but we won't make any direct use of that, because that will not provide any new information not already gathered by the $C_p$ statistic for a linear regression model. If you wanted to find the uncorrected AIC for a given model, you can use the `extractAIC` function.

```{r}
extractAIC(lm(lpsa ~ lcavol, data = prost_c12))
extractAIC(lm(lpsa ~ lcavol + lweight, data = prost_c12))
```

Note that:

- these results are fairly comparable to the bias-corrected AIC we built above, and
- the `extractAIC` and `AIC` functions look like they give very different results, but they really don't.

```{r}
AIC(lm(lpsa ~ lcavol, data = prost_c12))
AIC(lm(lpsa ~ lcavol + lweight, data = prost_c12))
```

But notice that the differences in AIC are the same, either way, comparing these two models:

```{r}
extractAIC(lm(lpsa ~ lcavol, data = prost_c12)) - extractAIC(lm(lpsa ~ lcavol + lweight, data = prost_c12))

AIC(lm(lpsa ~ lcavol, data = prost_c12)) - AIC(lm(lpsa ~ lcavol + lweight, data = prost_c12))
```

- AIC is only defined up to an additive constant.
- Since the difference between two models using either `AIC` or `extractAIC` is the same, this doesn't actually matter which one we use, so long as we use the same one consistently.


### My Usual Presentation

Usually, I just plot the three useful things (adjusted $R^2$, $C_p$ and $BIC$) when I'm working with best subsets.

```{r, fig.height = 7}
p1 / p2 / p3
```

## Which Subsets Look Best? (Tabulation)

```{r}
tibble(AdjR2 = which.max(rs_summ$adjr2),
       Cp = which.min(rs_summ$cp),
       BIC = which.min(rs_summ$bic))
```

Our candidate models appear to be models 3 (minimizes BIC), 6 (minimizes $C_p$) and 7 (maximizes adjusted $R^2$).

### Print the Coefficients of the Candidate Models

Model 3, minimizing BIC, uses `lcavol`, `lweight` and `svi`.

```{r}
coef(rs_mods, id = 3)
```

Model 6, minimizing $C_p$, uses `lcavol`, `lweight` and `svi`, to which it adds `age`, and two indicator variables related to our multicategorical variables, specifically, one for `bph = Low` and one for `gleason = 6`.

```{r}
coef(rs_mods, id = 6)
```

Model 7, maximizing adjusted $R^2$, adds `lcp` to Model 6.

```{r}
coef(rs_mods, id = 7)
```

### Rerunning our candidate models

Note that the models are **nested** because model 3 is a subset of the predictors in model 6, which includes a subset of the predictors in model 7. We'll also include an intercept-only model, and the full model containing all predictors we considered initially, just to show you what that would look like.

```{r}
m.int <- lm(lpsa ~ 1, data = prost_c12)
model3 <- lm(lpsa ~ lcavol + lweight + svi, data = prost_c12)
model6 <- lm(lpsa ~ lcavol + lweight + svi + age +
                 (bph == "Low") + (gleason == "6"), 
             data = prost_c12)
model7 <- lm(lpsa ~ lcavol + lweight + svi + age + lcp +
                 (bph == "Low") + (gleason == "6"), 
             data = prost_c12)
m.full <- lm(lpsa ~ lcavol + lweight + svi + 
              age + bph + gleason + lcp + pgg45, 
             data = prost_c12)
```

## Validation of Candidate Models

### 10-fold Cross-Validation of `model3`

Model 3 uses `lcavol`, `lweight` and `svi` to predict the `lpsa` outcome. Let's do 10-fold cross-validation of this modeling approach, and calculate key summary measures of fit, specifically the $R^2$, RMSE and MAE across the validation samples. We'll use the tools from the `caret` package we've seen in prior work.

```{r}
set.seed(43201)
train_c <- trainControl(method = "cv", number = 10)

model3_cv <- train(lpsa ~ lcavol + lweight + svi,
                   data = prost_c12, method = "lm",
                   trControl = train_c)

model3_cv
```

### 10-fold Cross-Validation of `model6`

As above, we'll do 10-fold cross-validation of `model6`.

```{r}
set.seed(43202)
train_c <- trainControl(method = "cv", number = 10)

model6_cv <- train(lpsa ~ lcavol + lweight + svi + age +
                       (bph == "Low") + (gleason == "6"),
                   data = prost_c12, method = "lm",
                   trControl = train_c)

model6_cv
```

This shows a lower RMSE, higher R-squared and lower MAE than we saw in Model 3, so Model 6 looks like the better choice. How about in comparison to Model 7?

### 10-fold Cross-Validation of `model7`

As above, we'll do 10-fold cross-validation of `model7`.

```{r}
set.seed(43203)
train_c <- trainControl(method = "cv", number = 10)

model7_cv <- train(lpsa ~ lcavol + lweight + svi + age +
                       lcp + (bph == "Low") + 
                       (gleason == "6"),
                   data = prost_c12, method = "lm",
                   trControl = train_c)

model7_cv
```

Model 7 looks better still, with a smaller RMSE and MAE than Model 6, and a higher R-squared. It looks like Model 7 shows best in our validation.

## What about Interaction Terms?

Suppose we consider for a moment a much smaller and less realistic problem. We want to use best subsets to identify a model out of a set of three predictors for `lpsa`: specifically `lcavol`, `age` and `svi`, but now we also want to consider the interaction of `svi` with `lcavol` as a potential addition. Remember that `svi` a factor with levels 0 and 1. We could simply add a numerical product term to our model, as follows.

```{r}
prost_c12 <- prost_c12 %>% 
    mutate(svixlcavol = as.numeric(svi) * lcavol)

rs_mod2 <- regsubsets(
    lpsa ~ lcavol + age + svi + svixlcavol,
    data = prost_c12, nvmax = 4, nbest = 1)

rs_sum2 <- summary(rs_mod2)

rs_sum2
```

In this case, best subsets identifies the interaction term as an attractive predictor at the very beginning, before it included the main effects that are associated with it. That's a problem. 

To resolve this, we could either:

1. Consider interaction terms outside of best subsets, and only after the selection of main effects.
2. Use another approach to deal with variable selection for interaction terms.

Best subsets is a very limited tool. While it's a better choice for some problems that a stepwise regression, it suffers from some of the same problems, primarily because the resulting models are too optimistic about their measures of fit quality.

Some other tools, including ridge regression, the lasso and elastic net approaches, are sometimes more appealing.

You may be wondering if there is a "best subsets" approach which can be applied to generalized linear models. There is, in fact, a `bestglm` package in R which can be helpful.

But, for now, it's time to move away from model selection and on to the very important consideration of non-linearity in the predictors.

