---
title: "431 Class 14"
author: "github.com/THOMASELOVE/2019-431"
date: "2019-10-10"
output:
  beamer_presentation:
    theme: "Madrid"
    fonttheme: "structurebold"
    colortheme: "whale"
    fig_height: 5.5
    fig_caption: false
---

```{r set-options, echo=FALSE, cache=FALSE, message = FALSE}
knitr::opts_chunk$set(comment=NA)
options(width = 55)
```

## Today's Agenda (Notes Chapters 19-20)

Statistical Inference and the `dm431` data

1. Paired vs. Independent Samples
2. Moving from Wide to Long and back again
3. Comparing Population Means using Independent Samples
    - Pooled t / Indicator Variable Regression
    - Welch's t 
    - Wilcoxon-Mann-Whitney rank sum
    - Bootstrap with `bootdif`

## Today's Setup and Data

```{r load_packages, message = FALSE}
library(magrittr); library(janitor) 
library(patchwork); library(here); 
library(boot); library(broom)
library(tidyverse)

source(here("R", "Love-boost.R"))

dm431 <- readRDS(here("data", "dm431.Rds"))
```

# Comparing Population Means

## So far, we've been thinking about one population, and one sample

Our population: ALL adults ages 31-70 seen for care this year and two years ago who live in Northeast Ohio with a diabetes diagnosis.

Our sample: 431 of those people, drawn in a way we hope is representative (but certainly isn't random).

## Are these Samples Paired (Matched) or Not?

Now, suppose we want to compare two subpopulations of our bigger population, using the relevant subsamples of our `dm431` data.

- Deciding whether or not the samples are paired (matched) is something we do before we analyze the data.

The best way to establish whether a study uses paired or independent samples is to look for the **link** between the two measurements that creates paired differences. 

## `dm431` Example 1.

Suppose we want to compare the mean `ldl` cholesterol level for subjects who are currently taking a `statin` medication to the mean `ldl` for subjects who are not currently taking a statin.

```{r}
dm431 %>% select(subject, ldl, statin) %>% tail()
```

## `dm431` Example 1.

Suppose we want to compare the mean `ldl` cholesterol level for subjects who are currently taking a `statin` medication to the mean `ldl` for subjects who are not currently taking a statin.

```{r, message = FALSE}
mosaic::favstats(ldl ~ statin, data = dm431) 
```

- What is the outcome of interest here? 
- What are the two exposure groups we are comparing?
- Does this design create paired samples or independent samples?

## `dm431` Example 2.

Suppose we want to compare the mean `ldl` cholesterol level for a set of subjects this year to the mean `ldl` for the same subjects two years ago.

```{r}
dm431 %>% select(subject, ldl, ldl_old) %>% head()
```

## `dm431` Example 2.

Suppose we want to compare the mean `ldl` cholesterol level for a set of subjects this year to the mean `ldl` for the same subjects two years ago.

```{r, message = FALSE}
mosaic::favstats(~ ldl, data = dm431) 
mosaic::favstats(~ ldl_old, data = dm431)
```

- What is the outcome of interest here? 
- What are the two exposure groups we are comparing?
- Does this design create paired samples or independent samples?

## `dm431` Example 3.

Suppose we want to compare the mean systolic blood pressure for male subjects to the mean systolic blood pressure among female subjects?

```{r}
dm431 %>% select(subject, sbp, sex) %>% head()
```

## `dm431` Example 3.

Suppose we want to compare the mean systolic blood pressure for male subjects to the mean systolic blood pressure among female subjects?

```{r, message = FALSE}
mosaic::favstats(sbp ~ sex, data = dm431) 
```

- What is the outcome of interest here? 
- What are the two exposure groups we are comparing?
- Does this design create paired samples or independent samples?

## Formatting the Data (Wide vs. Long)

**Wide** format (most appropriate for paired/matched samples) 

subject | treatment1 | treatment2
------: | ---: | ---:
A | 140 | 150
B | 135 | 145
C | 128 | 119

**Long** format (most appropriate for independent samples)

subject | sbp | group
------: | ---: | -----:
A | 140 | treatment1
A | 150 | treatment2
B | 135 | treatment1
B | 145 | treatment2
C | 128 | treatment1
C | 119 | treatment2

## Suppose you have a wide data set...

```{r}
tempdat_wide <- tibble(
  subject = c("A", "B", "C"),
  treatment_1 = c(140, 135, 128),
  treatment_2 = c(150, 145, 119)
)

tempdat_wide
```

## Pivot Data to make it longer

We want more rows, fewer columns. Each subject*treatment combination will become a row.

```{r}
tempdat_long <- tempdat_wide %>% 
  pivot_longer( -subject,
    names_to = "group", values_to = "sbp")
tempdat_long
```

## Spread the Data from Long to Wide

```{r}
tempdat_wide2 <- tempdat_long %>% 
  pivot_wider(names_from = group, values_from = sbp)

tempdat_wide2
```

## So, an independent samples design?

- Independent samples designs do not impose a matching, but instead sample two unrelated sets of subjects, where each group receives one of the two exposures. 
- The two groups of subjects are drawn independently from their separate populations of interest. 
- One obvious way to tell if we have an independent samples design is that this design does not require the sizes of the two exposure groups to be equal.

## Three Small Analyses using Independent Samples

Our population: ALL adults ages 31-70 seen for care this year and two years ago who live in Northeast Ohio with a diabetes diagnosis.

Our sample: 431 of those people, drawn in a way we hope is representative (but certainly isn't random).

1. Can we estimate the difference in the population mean systolic blood pressure among females in our population as compared to males in our population?

2. Can we estimate the difference in the population mean LDL level for those on a statin as compared to those not on a statin?

3. Can we estimate the difference in the population mean hemoglobin A1c for those with Medicaid vs. Medicare insurance?

## SBP for females vs. males, in a boxplot

```{r, echo = FALSE}
ggplot(dm431, aes(x = sex, y = sbp)) +
    geom_violin() +
    geom_boxplot(aes(fill = sex), width = 0.3, notch = TRUE) +
    coord_flip() +
    guides(fill = FALSE) +
    theme_bw() + 
    labs(x = "", y = "Systolic Blood Pressure this year",
         title = "Independent Samples Comparison: SBP by Sex")
```

## Another Picture for Two Independent Samples

```{r, echo = FALSE}
ggplot(dm431, aes(x = sbp, fill = sex)) +
  geom_histogram(bins = 12, col = "white") +
  facet_wrap(~ sex) +
  guides(fill = FALSE) + 
  labs(title = "Systolic Blood Pressure by Sex in 431 Adults with Diabetes")
```

## Numerical Summary for Two Independent Samples

```{r}
mosaic::favstats(sbp ~ sex, data = dm431)
```

## Independent Samples: Confidence Intervals for $\mu_1 - \mu_2$

1. Pooled t CI or Indicator Variable Regression Model (t approach assuming equal population variances)
2. Welch t CI (t approach without assuming equal population variances)
3. Wilcoxon-Mann-Whitney Rank Sum Test (non-parametric test not assuming Normality but needing symmetry to be related to means)
4. Bootstrap confidence interval for the difference in population means (fewest assumptions of these options)

## Results for the SBP and Sex Study

Procedure     | *p* for $H_0: \mu_F = \mu_M$ | 95% CI for $\mu_F - \mu_M$
:-----------: | --------------------: | :------------------------:
Pooled t  | 0.90 | (-3.3, 3.8)
Welch t   | 0.89 | (-3.2, 3.7)
Bootstrap   | *p* > 0.05 | (-3.1, 3.6)

Procedure     | *p* for $H_0: psmed_F = psmed_M$ | 95% CI for M - F shift
:-----------: | --------------------: | :------------------------:
Rank Sum  | 0.42 | (-2.0, 5.0)

What conclusions should we draw, at $\alpha$ = 0.05?

