---
title: "Study A Survey Data After Merging"
author: "Thomas E. Love"
date: "2019-11-22"
output: pdf_document
---

# Steps to use the data I provided

1. Create a directory called `431projectA` on your computer.
2. Create a subdirectory called `studyAdata` on your computer, underneath `431projectA.`
3. Place the five data sets in your `studyAdata` directory.
4. Create a subdirectory called `R` on your computer, also underneath `431projectA`.
5. Place the `Love-boost.R` file in the `R` subdirectory you just created.
6. Open R Studio
7. Create a new R project in your `431projectA` directory using **File ... New Project** and then select Existing Directory, and feed it your `431projectA` main directory. That is where your R Markdown files will live.

With this setup in mind, I then used the following code to obtain the results below.

# Initial Setup

```{r setup}
knitr::opts_chunk$set(comment = NA)
```

```{r}
library(magrittr); library(here); library(janitor)
library(readxl); library(tidyverse)
```

```{r}
studyA_file1 <- read_csv(here("studyAdata", "studyA-2019-student-data-01.csv")) %>%
    mutate(subject = as.character(subject)) 

studyA_file2 <- read_xls(here("studyAdata", "studyA-2019-student-data-02.xls")) %>%
    mutate(subject = as.character(subject)) %>%
    tbl_df

studyA_file3 <- read_xls(here("studyAdata", "studyA-2019-student-data-03.xls")) %>%
    mutate(subject = as.character(subject)) %>%
    tbl_df

studyA_file4 <- read_csv(here("studyAdata", "studyA-2019-student-data-02.csv")) %>%
    mutate(subject = as.character(subject))

studyA_file5 <- read_csv(here("studyAdata", "studyA-2019-student-data-02.csv")) %>%
    mutate(subject = as.character(subject))

```



# `Hmisc::describe` for the entire data set

```{r}
Hmisc::describe(studyA)
```

# Check All That Apply items

```{r}
studyA %>% tabyl(Q029)
```

```{r}
studyA %>% tabyl(Q030)
```

```{r}
studyA %>% tabyl(Q055)
```