---
title: "Study A Survey Data After Merging with HINTS"
author: "Thomas E. Love"
date: "`r Sys.Date()`"
output: pdf_document
---

# Steps to use the data I provided

1. Create a directory called `431projectA` on your computer.
2. Create a subdirectory called `studyAdata` on your computer, underneath `431projectA.`
3. Place the five data sets in your `studyAdata` directory.
    - Be sure to use the revised data sets posted on 2019-11-22 to 
4. Create a subdirectory called `R` on your computer, also underneath `431projectA`.
5. Place the `Love-boost.R` file in the `R` subdirectory you just created.
6. Open R Studio
7. Create a new R project in your `431projectA` directory using **File ... New Project** and then select Existing Directory, and feed it your `431projectA` main directory. That is where your R Markdown files will live.

With this setup in mind, I then used the following code to obtain the results below.

# Initial Setup

```{r setup}
knitr::opts_chunk$set(comment = NA)
```

```{r, message = FALSE}
library(magrittr); library(here); library(janitor)
library(readxl); library(tidyverse)
```

## Read in the data files

```{r, message = FALSE}
studyA_raw1 <- read_csv(here("studyAdata", "studyA-2019-student-data-01.csv")) 

studyA_raw2 <- read_xls(here("studyAdata", "studyA-2019-student-data-02.xls")) 

studyA_raw3 <- read_xls(here("studyAdata", "studyA-2019-student-data-03.xls")) 

studyA_raw4 <- read_csv(here("studyAdata", "studyA-2019-student-data-04.csv")) 

studyA_raw5 <- read_xls(here("studyAdata", "studyA-2019-student-data-05.xls")) 
```

## Combine raw2 and raw3

```{r}
studyA_raw23 <- bind_rows(studyA_raw2, studyA_raw3)

dim(studyA_raw23)
```

## Combine raw4 and raw5

```{r}
studyA_raw45 <- bind_rows(studyA_raw4, studyA_raw5)

dim(studyA_raw45)
```

## Merge raw1 with raw23

```{r}
studyA_raw123 <- inner_join(studyA_raw1, studyA_raw23, by = "subject")

dim(studyA_raw123)
```

## Merge raw123 with raw45

```{r}
studyA_raw12345 <- inner_join(studyA_raw123, studyA_raw45, by = "subject")

dim(studyA_raw12345)
```

## Change character variables to factors throughout, except "subject"

```{r}
studyA_clean <- studyA_raw12345 %>%
    mutate_if(is.character, as.factor) %>%
    mutate(subject = as.character(subject))

studyA_clean
```

# `Hmisc::describe` for the entire data set

```{r}
Hmisc::describe(studyA_clean)
```


# `tabyl`s for the "Check All That Apply" items

```{r}
studyA_clean %>% tabyl(Q029)
```

```{r}
studyA_clean %>% tabyl(Q030)
```

```{r}
studyA_clean %>% tabyl(Q055)
```

Good luck!